Efficient implementation of penalized models and their application to genetic risk prediction

### 2018-05-10 paper/draft.tex
modif discussion

In this comparative study, we propose penalized logistic regressions as a replacement for the widely used C+T method for computing polygenic risk scores when individual genotype data are available. We show that jointly estimating SNP effects using efficient penalized logistic regressions can achieve higher predictive performance, as reported in \cite{abraham2013performance}. Indeed, when disease architecture and sample size allow for high predictive performance (say AUC greater than 85\%), ``logit-simple'' provides higher predictive performance than the widely-used C+T method, nearly reaching optimal AUC values, which is encouraging for using polygenic risk scores in clinical settings.
The C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not yet the case for our penalized logistic regressions. However, in order to accurately predict disease status for clinical interventions, best possible predictive methods should be used. The penalized logistic regression we present in this paper could be used on data as large as biobank datasets because it is both very memory- and computation-efficient.

As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the method \cite[]{ware2017heterogeneity}. On the contrary, in the penalized logistic regression presented here, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance [TODO: COMPARAISON AVEC LE MAX DE BIGLASSO]. Usually, cross-validation is used to assess the best value for the hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}. Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we prefer to use CMSA when there is a model to choose from a regularization path, with the possibility of adding an early stopping criterion to fasten the whole procedure. We also introduce a simple feature engineering technique which enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive ones. This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects. Moreover, it also improves predictive performance for the celiac disease.Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}. In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population. We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}. In another study, we wish also to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.

>>>>

\subsection{Joint estimation improves predictive performance}

In this comparative study, we present a computationally efficient implementation of penalized logistic regression. This model can be used to build polygenic risk scores based on very large SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to dramatically improve predictive performance as compared to the standard C+T approach in which these effects are learned independently. Penalized logistic regression consistently outperform the C+T method, excepted in the case where there are only small effects and sample size is limited, in which case all methods perform poorly. Note that in this particular case, penalized logistic regression could outperform the C+T method if choosing a different value for hyper-parameter $\alpha$.

\subsection{Importance of hyper-parameters}

Indeed, the choice of hyper-parameters' values is very important as it can greatly impact the performance of the methods. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
Moreover, in this paper, we reported the maximum AUC of 102 $p_T$ thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method \cite[]{ware2017heterogeneity}, which we confirm in this study.

On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing an optimal hyper-parameter $\lambda$, accouting for the amount of regularization used in the model. Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to assess the best value for this hyper-parameter and then the model is train again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}. Yet, as the value of $\lambda$ depends on the sample size and doing cross-validation and retraining is computationally demanding, we think using CMSA is more appropriate in this setting. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we show that an hybrid approach ($\alpha = 0.5$) gives similar results as a model using only L1 regularization ($\alpha = 1$). 
Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, when testing with $\alpha = 0.001$, we show that this setting consistently provides lower predictive performance, excepted in the case where there are many small effects. In the light of this result, we now integrate an optional grid-search procedure in our implementation in order to determine an optimal value for hyper-parameter $\alpha$.

\subsection{Non-linear effects}

In this paper, we also explored how to capture non-linear effects. For this, we introduced a simple feature engineering technique that enables the logistic regression to detect and learn not only additive effects, but also dominant and recessive effects. This technique improves the predictive performance of the method when there are some non-linear effects in the simulations, while providing the same predictive performance when there are only linear effects. Moreover, it also improves predictive performance for the celiac disease.

Yet, this approach is not able to detect interaction-type effects. 
In order to capture interaction-type effects, we tested T-Trees, a method that is able to exploit SNP correlations thanks to special decision trees \cite[]{botta2014exploiting}. 
However, predictive performance of T-Trees were consistently lower than penalized logistic regression, even when simulating a fancy model in which there are dominant and interaction-type effects that T-Trees should be able to handle.

\subsection{Limitations}

Our approach has one major limitations: the main advantage of the C+T method is that it is applicable directly on summary statistics, allowing to leverage the largest GWAS sample size to date, even when individual cohort data cannot be merged because of practical and ethical reasons (e.g. consortium data including hundreds of cohorts). As of today, the proposed penalized logistic regression does not allow for the analysis of summary data, but represents an important future direction of our work. The current version is of particular interest for the analysis of modern SNP dataset including hundreds of thousands of individuals.

Finally, in this comparative study, we did not consider the problem of population structure \cite[]{marquez2017multiethnic,martin2017human,vilhjalmsson2015modeling} and also did not consider non-genetic data such as environmental and clinical data \cite[]{dey2013integration,van2012integration}. In another study, we will assess how can we use models and effects learned in one population to improve learning and prediction in another population. We will use a dataset as large as the UK biobank \cite[]{bycroft2017genome}. In another study, we wish also to assess how can we combine the information provided by genetic data with clinical and environmental data, possibly in a non-linear way.
